Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-07 paper/paper.tex
Introduction

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or
productivity. For recursion, this is done by automatically inferring
which argument in the recursive call gets smaller with respects to the original
input argument. For productivity, the algorithm checks that the corecursive
call appears directly under a constructor to make sure that this function
always produces at least one element after each recursive step.
This implies that some functions, though well-defined, cannot be accepted by
the proof assistant. Examples of this include common sorting algorithms, such
as quicksort:
%One example is Quicksort, written in OCaml
%below:
\begin{minted}{OCaml}
let rec qsort xs = match divide xs with | None -> []
 | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant. The reason is that the \ocaml{divide}
function \emph{destructuring} the input dives deeper in the input returning
two sublists with the head as a pivot.

The main approach for implementing non-structural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are accompanied by
\emph{termination proofs}. Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
one approach is using the \coq{Fix} combinator, in which recursion is done on an
\emph{accesibility} predicate on the input, which is a proof that there are no
infinitely decreasing chains.}.
A benefit of this strategy is that when extracting verified non-structural
recursive functions to OCaml, the additional proofs are discarded, resulting in
code that closely resembles what a programmer might have written directly in
OCaml. However, reasoning about program equivalences needs  such termination
proofs to be kept around, leading to the common practice of employing additional
reduction lemmas extensively.

% A benefit of this approach is that, extracting
% verified nonstructural recursive functions to OCaml will erase the additional
% proofs, thus producing code that will be closer to what a programmer may have
% written directly in OCaml. However, reasoning about program equivalences
% requires dealing with such termination proofs, and it is common practice to use
% custom reduction lemmas that will be used extensively when proving properties
% about nonstructural recursive functions.

\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
for structuring recursive definitions that has found a lot of success in
functional programming. The idea is that, rather than writing a self-refential
definition, we use a function library, the recursion scheme, which takes as
input the non-recursive parts of the algorithm and returns the desired recursive
definition which is guaranteed to terminate by the properties of the recursion
scheme itself. More complex recursion schemes do exist but it has been
proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
This result, thus, yields a general framework to reason about the algebraic properties of these
algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.

@@

However, when used in the context of languages with general recursion like
(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is
because the input information itself could be non-terminating. On the other
hand, while formalising recursion schemes in a type theory does provide stronger
termination guarantees, to the best of our knowledge, not many attempts have
been made at mechanising them.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, in Abreu et al.'s approach, extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
programmers still need to reason about the termination of their programs, our
mechanisation will allow the use of the algebraic laws of hylomorphisms for
program reasoning, as well as extracting to idiomatic code (see
extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

To summarise, our contributions are as follows. In this paper we provide a
framework for \emph{generic programming} with recursion schemes in Coq, with a
proof of their algebraic laws and program equivalences for clean code
extraction:
\begin{itemize}
  \item In Section~\ref{sec:containers} we formalise the type of container functors ensuring the presence of least and greatest fixed-points for functors and suitably adapted for program extraction
  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and hylomorhisms and their universal properties
  \item In Section~\ref{sec:examples} we use the framework to formalise examples of divide-and-conquer, dynamic programming, and mutual recursion algorithms. Furthermore we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:intro}
Structured recursion schemes~\cite{HinzeW16, HinzeWG15} are powerful
abstractions that capture common patterns of recursion. The main benefit of
structuring computation using recursion schemes, is that they enjoy
well-estabished \emph{algebraic properties} that can serve as a foundation for
program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}).  These
algebraic properties led to their use in the context of \emph{program
calculation}, where programmers would describe their code using simple,
inefficient specifications in an \emph{algebra of programming}~\cite{}, and then
use the algebraic laws of this algebra of programming to \emph{calculate} an
efficient version of the same algorithm.

\dcas{TODO: Introduce example}

Despite the rigorous algebraic laws that are satisfied by recursion schemes,
there is a lack of tool support for their use in the context of program
calculation. In fact, most of the work on applying program calculation is done
by performing pen-and-paper proofs, and then translating the result to specific
instances of recursion schemes, implemented in a programming language (generally
Haskell)~\cite{TODO:INeedToFindSomeMoreReferences}. Most of the work on
automating program calculation techniques either build implementations of the
algebraic laws, together with some form of syntax or
EDSL~\cite{TODO:INeedToFindSomeMoreReferences}. Such implementations are not
verified, and can introduce bugs in the way they apply the algebraic laws as
rewriting rules.

A more rigorous approach to implementing program calculation techniques is to
\emph{mechanise} them as part of a proof assistant. However, most of the work in
mechanising recursion schemes focuses on a narrow subset of the known recursion
schemes due to termination
issues~\cite{TODO:FillInReferencesForCoqMechanisationsFromRelatedWorkSection}.
Indeed, most of the mechanisations focus on \emph{maps} and \emph{folds}, and
few authors focus on the \emph{unfolds}. In general, Hinze et
al.~\cite{HinzeWG15} has proven that all known  recursion schemes are instances
of a  divide-and-conquer algorithm which goes under the name of
\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To date, no work has mechanised
hylomorphisms in the Coq proof assistant.

%Recursive definitions cannot be proven well-defined automatically due to the
%halting problem. Modern proof assistants like Coq or Agda provide a sound, but
%incomplete algorithm which syntactically checks for termination or
%productivity. For recursion, this is done by automatically inferring
%which argument in the recursive call gets smaller with respects to the original
%input argument. For productivity, the algorithm checks that the corecursive
%call appears directly under a constructor to make sure that this function
%always produces at least one element after each recursive step.
%This implies that some functions, though well-defined, cannot be accepted by
%the proof assistant. Examples of this include common sorting algorithms, such
%as quicksort:
%%One example is Quicksort, written in OCaml
%%below:
%\begin{minted}{OCaml}
%let rec qsort xs = match divide xs with | None -> []
% | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
%\end{minted}
%While \ocaml{qsort} is a well-defined mathematical function it cannot be
%accepted by a proof assistant. The reason is that the \ocaml{divide}
%function \emph{destructuring} the input dives deeper in the input returning
%two sublists with the head as a pivot.
%
%The main approach for implementing non-structural recursion in Coq is to use
%\emph{well-founded recursion}, where recursive definitions are accompanied by
%\emph{termination proofs}. Using well-founded recursion, the recursive calls
%will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
%one approach is using the \coq{Fix} combinator, in which recursion is done on an
%\emph{accesibility} predicate on the input, which is a proof that there are no
%infinitely decreasing chains.}.
%A benefit of this strategy is that when extracting verified non-structural
%recursive functions to OCaml, the additional proofs are discarded, resulting in
%code that closely resembles what a programmer might have written directly in
%OCaml. However, reasoning about program equivalences needs  such termination
%proofs to be kept around, leading to the common practice of employing additional
%reduction lemmas extensively.
%
%% A benefit of this approach is that, extracting
%% verified nonstructural recursive functions to OCaml will erase the additional
%% proofs, thus producing code that will be closer to what a programmer may have
%% written directly in OCaml. However, reasoning about program equivalences
%% requires dealing with such termination proofs, and it is common practice to use
%% custom reduction lemmas that will be used extensively when proving properties
%% about nonstructural recursive functions.

%\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
%for structuring recursive definitions that has found a lot of success in
%functional programming. The idea is that, rather than writing a self-refential
%definition, we use a function library, the recursion scheme, which takes as
%input the non-recursive parts of the algorithm and returns the desired recursive
%definition which is guaranteed to terminate by the properties of the recursion
%scheme itself. More complex recursion schemes do exist but it has been
%proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
%algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
%This result, thus, yields a general framework to reason about the algebraic properties of these
%algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.

@@

%However, when used in the context of languages with general recursion like
%(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is
%because the input information itself could be non-terminating. On the other
%hand, while formalising recursion schemes in a type theory does provide stronger
%termination guarantees, to the best of our knowledge, not many attempts have
%been made at mechanising them.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration is what allows the execution of code that has been verified in
Coq~\cite{OnoHTNH11, Larchey-Wendling23, MiculanP12, Sakaguchi20}.  In Abreu et
al.'s approach, extraction (1) do not preserve the recursive structure of common
implementations; and (2) lead to unsafe casts like \ocaml{Obj.magic} in the
generated code. This latter is also problematic in that, for higher-order
programs, simple interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
programmers still need to reason about the termination of their programs, our
mechanisation will allow the use of the algebraic laws of hylomorphisms for
\emph{program calculation}, as well as extracting to idiomatic code (see
extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

To summarise, our contributions are as follows. In this paper we provide a
framework for \emph{program calculation} with recursion schemes in Coq, where
program calculation is done by applying regular Coq equalities, based on  proofs
of the algebraic laws of hylomorphisms, and that allows the extraction of safe,
idiomatic OCaml code:
\begin{itemize}
  \item In Section~\ref{sec:containers} we formalise the type of container functors ensuring the presence of least and greatest fixed-points for functors and suitably adapted for program extraction
  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and hylomorhisms and their universal properties
  \item In Section~\ref{sec:examples} we use the framework to formalise examples of divide-and-conquer, dynamic programming, and mutual recursion algorithms. Furthermore we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
\end{itemize}
